---
title: "An Introducation to R"
author: "Dean Adams (dcadams@iastate.edu)"
date: "`r Sys.Date()`"
output: html_document
---

### **Motivation**
R is a statistical programming environment with many built-in mathematical functions and many others that are found in packages that can be installed. `RRPP` is one such package.  

Analyses in R are performed using a series of commands which are written in script files and passed to the main console. The general workflow is:

* 1: OPEN R
* 2: OPEN R-script
* 3: change to working directory 
* 4: run analyses

#### **1: Some basic commands**
Like any programming language, one must learn its syntax. Over time one learns the commands in R, and how to string them together into meaningful operations. Here are some basic commands:

```{r eval=TRUE,error=TRUE}
3->a  #Assign value to variable
a

b<-c(3,4,5)   # 'c' combines values into vector or list
b
b[2]	#access items in vectors by calling their position

a <- rnorm(50) #generate random normal vector
b <- rnorm(a)
plot(a, b)      # a simple plot

c<-cbind(a,b)    #binds columns together (rbind does same by rows)
c
c[1]   #first element
c[1,]	 #first row
c[,1]  #first column

rbind(a,b)

ls() # See which R objects are now in the R workspace.

### Some base functions
sum(a)
mean(a)
min(a)
max(a)
var(a)
a^2		#square values
sqrt(a)	#NaN for negative values
abs(a)
cor(a,b)

rm(list=ls())   #remove items in memory

# Matrix operations
a<-matrix(c(1,0,4,2,-1,1),nrow=3)
b<-matrix(c(1,-1,2,1,1,0),nrow=2)
a
b

c<-t(a)	#matrix transpose
a
c

2*a	#scalar multiplication

#Matrix addition and subtraction
b+c
b-c
a+b  ##NOTE: non-conformable matrices (check rxc of your matrices!)

#elementwise multiplication (hadamard product)
c
b
c*b

# matrix multiplication
a%*%b		## %*% is symbol for matrix multiplication
b%*%a		## matrix order matters
	
rm(list=ls())


gl(2,10)
```

#### **2: Reading Data**
There are many ways to read data into R. Here is one example:

```{r eval=TRUE}
mydata<-read.csv(file="Data/Lab-01-RIntroData.csv",header=T)
mydata
Y<-as.matrix(mydata[,(2:3)])
FactorA<-as.factor(mydata[,4])
Y
FactorA

```

#### **3: Other Functions: the apply family**

A useful set of functions are the 'apply' family of functions. These perform some task repetitively in some way over a data structure. A few simple examples follow:

```{r eval=TRUE,error=TRUE}
apply(Y,2,sd)    #here, we obtain the std for each column of a matrix

tapply(Y[,1],FactorA,mean)	#Obtain means for first column for levels of FactorA
tapply(Y[,2],FactorA,mean)	#Obtain means for first column for levels of FactorA
tapply(Y,FactorA,mean)		  #Try entire matrix: doesn't work

rowsum(Y, FactorA)/as.vector(table(FactorA))
```

#### **4: Building Functions**

One can also make your own function. These are defined as function(xx)  where `xx` defines the input types (there may be more than one). Below is a simple example (NOTE: a good way to learn the syntax of functions is to type the name of the function at the command line without (). This will then show the underlying code of the function):

```{r eval=TRUE}
mymean<-function(x){
  n<-length(x)
  tmp<-0
  for (i in 1:n){
    tmp<-tmp+x[i]
  }
  mn<-tmp/n
  return(mn)
}

x<-rnorm(10)
mean(x)  
mymean(x)  #works!
```

#### **5: Basic Statistical Models**

Because R is a *statistical* programming language there are many statistical models in R.  Here are a few basic ones (we'll discover many more this semester):

```{r eval=TRUE}
model1<-lm(mydata$y~mydata$x)  #run regression
summary(model1)
anova(model1)  #generates anova table of results

#A plot with regression line
plot(mydata$x,mydata$y)
abline(coef(model1))

model2<-lm(mydata$y~mydata$groups)  #run anova
summary(model2)
anova(model2)
```